---
title: "RepData_PeerAssessment1"
author: "Pradeept Sahoo"
date: "March 3, 2018"
output: html_document
---

# Reproducible Research : Peer Assignment 1

## Loading and Processing the Data  
To load the data I use the read.csv() command.
1.Code for reading in the dataset and/or processing the data
```{r loaddata}
setwd("C:/Coursera/R/ReprodResearch")
activityData <- read.csv("activity.csv")
```

To see what the data look like, I use the str(), summary(), and head() commands:  

```{r checkdata}
str(activityData)
summary(activityData)
head(activityData)
```

Modify the Date String to Date class and Remove the missing values and create a new data set. 


```{r finalData}
# change date type to date class 
activityData$date <- as.Date(as.character(activityData$date))
# create NA dataset
activityDataNA <- is.na(activityData$steps)
finalActivityData <- activityData[!activityDataNA,]
head(finalActivityData)
```

Calculate the total number of steps taken per day
```{r stepsPerDay}
acivityByDaySum <- aggregate(finalActivityData$steps, by=list(finalActivityData$date),sum)
names(acivityByDaySum)[1] = "date"
names(acivityByDaySum)[2] = "totalsteps"
head(acivityByDaySum,10)
```

2.Histogram of the total number of steps taken each day

```{r stepsPerDayHistogram}
# plot using ggplot
library(ggplot2)
ggplot(acivityByDaySum, aes(x=totalsteps)) +
      geom_histogram(fill="steelblue",  bindwidth=1000) + 
      labs(title="Total Daily Steps", x="Steps", y="Frequency")
```

3.Mean and median number of steps taken each day
```{r tstepsMeanMedian}
mean(acivityByDaySum$totalsteps)
median(acivityByDaySum$totalsteps)
```

## Average daily activity pattern
4.Time series plot of the 5 minutes interval and the average number of steps taken

```{r timeSeries}
# subset by interval
meanDataByInterval <- aggregate(finalActivityData$steps, by=list(finalActivityData$interval), mean)
# set the column names
names(meanDataByInterval)[1]="interval"
names(meanDataByInterval)[2]="steps"
ggplot(meanDataByInterval, aes(x=interval, y=steps))+
  labs(title = "Sum of the Steps by Interval", x="interval", y="steps")+
  geom_line(color="red")

```

5.The 5-minute interval, on average across all the days that contains the maximum number of steps

```{r maxSteps}
maxIntervalSteps <- meanDataByInterval[which.max(meanDataByInterval$steps),]
maxIntervalSteps
```
6.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missingValues}
missingValues <- sum(activityDataNA)
missingValues
```
7. Make a histogram of the total number of steps taken each day after missing values are imputed

```{r}
# new dataset with NA values
activityData2 <- activityData
# Remove NA values
cleanDataSet <- activityData2[!is.na(activityData2),]
 
# Date wise sum data
summedActivityByDay <- aggregate(cleanDataSet$steps, by = list(cleanDataSet$date),sum)
names(summedActivityByDay)[1]="date"
names(summedActivityByDay)[2]="totalsteps"
## Create Histogram using ggplot2
ggplot(summedActivityByDay, aes(x=totalsteps))+
  geom_histogram(fill="steelblue", binwidth=1000) +
  labs(title="Total Daily Steps", x="Steps", y= "Frequency")
```
8. Differences in activity patterns between weekdays and weekends. 
```{r weekvsweekend}
baseData2 <- cleanDataSet
baseData2$weekday <- weekdays(baseData2$date)
baseData2$weekend <- ifelse (baseData2$weekday == "Saturday" | baseData2$weekday == "Sunday", "Weekend", "Weekday")

MeanDataWeekendWeekday <- aggregate(baseData2$steps, by=list(baseData2$weekend, baseData2$interval), mean)
names(MeanDataWeekendWeekday)[1] ="weekend"
names(MeanDataWeekendWeekday)[2] ="interval"
names(MeanDataWeekendWeekday)[3] ="steps"
#head(MeanDataWeekendWeekday,10)
ggplot(MeanDataWeekendWeekday, aes(x = interval, y=steps, color=weekend)) +
  geom_line() +
  facet_grid(weekend ~ .) +
  labs(title = "Mean of Steps by Interval", x = "interval", y = "steps")  
    
```

There seems to be an overall slightly larger incidence of steps during the weekends.  
